Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices

### 2022-11-01 ms/ms/ms.Rmd
<edit ms>

---
title: |
  Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices
type: ARTICLE TEMPLATE
author:
  - name: Robert Z. Selden, Jr.
    affil: a
    email: zselden@sfasu.edu
  - name: John E. Dockall
    affil: b
    email: john.dockall@stantec.com
  - name: David H. Dye
    affil: c
    email: daviddye@memphis.edu
affiliation:
  - num: a
    address: |
      Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; and Cultural Heritage Department, Jean Monnet University 
  - num: b
    address: |
      Stantec, Inc.
  - num: c
    address: |
      Department of Earth Sciences, The University of Memphis
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  Elliptical bifaces are prevalent at Mississippian sites throughout the American Southeast. Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area. Elliptical bifaces from Millsap Cache were produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material. The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar—convergent—biface shape in the small class. Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This suggests that a shared, and consistent, mental template was maintained independent of biface size. The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does. This finding supports the argument that elliptical biface morphology diverges through local reduction practices. As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.
keywords: |
  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; lithics; oblique parallel flaking; archaeoinformatics; museum studies; digital humanities; non-western art history; geometric morphometrics; STEM; STEAM
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/elliptical.illustration.png}
\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production. The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region.}
\label{fig:illustrated}
\end{figure}

# Methods

To assess the variable impacts that Caddo retouch practices may have had on general morphology, the sample of elliptical bifaces was binned into two size classes using centroid size. Due to potential morphological differences associated with the use of two distinct raw materials (Kay County flint and Edwards chert), the sample was subset by site prior to calculating the mean centroid size for each. Bifaces larger than the mean were then assigned to the large category, those smaller than the mean were assigned to the small category, and the two datasets were then joined in advance of analysis ([Supplementary Materials, Ch. 1](https://seldenlab.github.io/elliptical.bifaces/define-size-classes.html)).

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/ellipticalbifaces.png}
\caption{Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the flaking index developed and advanced by \citet{RN11099}, \citet{RN9242}, and \citet{RN11098}.}
\label{fig:elliptical}
\end{figure}

To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by \citet{RN11099}, and later advanced and refined by \citet{RN9242}, then \citet{RN11098}. The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length inclusive of curvature, which was calculated using ImageJ (Abramoff et al. 2004, Rasband 1997, Schneider et al. 2012). The number of flake scars was then divided by the length of the worked edge, with the most heavily worked edge identified by the greater value.

@@

The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \ref{fig:elliptical}). Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R 4.2.1 (R Core Development Team, 2022) and the StereoMorph package (Olsen and Westneat 2015).

Landmark data were aligned to a global coordinate system (Bookstein et al. 1999, Gunz et al. 2005, Kendall 1981, 1984, Slice 2001), achieved through generalised Procrustes superimposition (Bookstein 1986, Rohlf and Slice 1990, Rohlf 1999) in R using the geomorph (Adams et al. 2017, Adams and Otarola-Castillo 2013, Baken et al. 2021) and RRPP packages (Adams and Collyer 2015, Collyer and Adams 2018). Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size—the square root of the sum of squared distances from each landmark to the specimen’s centroid (Chapman 1990, Dryden and Mardia 1998, Gower 1975, Rohlf and Slice 1990). The geomorph package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Dryden and Mardia 1993, Kent and Mardia 2001, Rohlf 1999, Slice 2001).

@@

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074). Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by support from the Heritage Research Center at Stephen F. Austin State University. A Research Support Fund Grant from the Texas Archeological Society provided the tool needed to illustrate the bifaces.

>>>>

---
title: |
  Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices
type: ORIGINAL RESEARCH ARTICLE
author:
  - name: Robert Z. Selden, Jr.
    affil: a
    email: zselden@sfasu.edu
  - name: John E. Dockall
    affil: b
    email: john.dockall@stantec.com
  - name: and David H. Dye
    affil: c
    email: daviddye@memphis.edu
affiliation:
  - num: a
    address: |
      Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; and Cultural Heritage Department, Jean Monnet University 
  - num: b
    address: |
      Stantec, Inc.
  - num: c
    address: |
      Department of Earth Sciences, The University of Memphis
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast. Those from Millsap Cache and Jowell Farm comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area. Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material. The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class. Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This suggests that a shared, and consistent, mental template was maintained independent of biface size. The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does. This finding supports the argument that elliptical biface morphology diverges through local reduction practices. As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.
keywords: |
  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; Jowell biface; lithics; oblique parallel flaking; archaeoinformatics; museum studies; digital humanities; non-western art history; geometric morphometrics; STEM; STEAM
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

# Methods and results

To assess the variable impacts that Caddo retouch practices may have had on general morphology, the sample of elliptical bifaces was first binned into two size classes using centroid size ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Due to potential morphological differences associated with the use of two distinct raw materials (Kay County flint and Edwards chert), the sample was subset by site prior to calculating the mean centroid size for each. Raw material for bifaces from Millsap Cache were identified by \citet{RN11461} as Kay County flint, and those from Jowell Farm were identified as Edwards Chert due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources. Bifaces larger than the mean were then assigned to the large category, those smaller than the mean were assigned to the small category, and the two datasets were then joined in advance of analysis ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/ellipticalbifaces.png}
\caption{Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area at top right identified using a modified version of the flaking index developed and advanced by \citet{RN11099}, \citet{RN9242}, and \citet{RN11098}.}
\label{fig:elliptical}
\end{figure}

To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by \citet{RN11099}, and later advanced and refined by \citet{RN9242}, then \citet{RN11098}. The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length inclusive of curvature, which was calculated using ImageJ \citep{RN11146,RN11147,RN11148}. The number of flake scars was then divided by the length of the worked edge, with the most heavily worked edge identified by the greater value.

@@

The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \@ref(fig:elliptical)). Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R \citep{R} and the `StereoMorph` package \citep{RN9091}.

### Generalised Procrustes Analysis

Landmark data were aligned to a global coordinate system \citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R using the `geomorph` and `RRPP` packages \citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid \citep{RN11139,RN11140,RN11564,RN478}. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Figure \ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/gpa.png}
\caption{Results of generalised Procrustes analysis for elliptical bifaces from the Millsap Cache and Jowell Farm. Mean consensus configuration shown in black; individual bifaces in gray.}
\label{fig:gpa}
\end{figure}

### Principal Components Analysis

Principal components analysis \citep{RN1746} was used to visualise shape variation among the elliptical bifaces, and the scatterplot and convex hulls represent the dispersion of shapes in tangent space (Figure \ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}. The shape space described by each principal axis is commonly visualized using thin-plate spline warping of a reference image \citep{RN1731,RN479}.

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/pca.png}
\caption{PCA summarizing shape variation in elliptical bifaces by site and size class. }
\label{fig:pca}
\end{figure}

### Procrustes ANOVA

To assess whether shape changes with size (allometry), and whether shape and size differed by raw material and size class, Procrustes ANOVAs \citep{RN1749} were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions \citep{RN1756}. A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs \citep{RN1655,RN11775}, which has higher statistical power and a greater ability to identify patterns in the data should they be present \citep{RN1719}.

The test for allometry in the sample did not return a significant result, suggesting that elliptical biface shape and size may follow a pattern of isometry, where shape remains constant through the reduction process ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Elliptical bifaces were also found to express a consistency of shape across all analyses ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Results demonstrated a significant difference in size by site/raw material (RRPP = 10,000, Pr(>F) = 0.0007); however, the subsequent analysis of elliptical bifaces by size class demonstrates that the pattern is more readily attributable to reduction practices/intensity rather than nodule and raw material size ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).

The analysis of morphology by size class demonstrates a significant difference in size between size classes; however, that pattern deviated from our expectations. Elliptical bifaces from Jowell Farm differ in neither shape nor size between the size classes ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Elliptical bifaces in the large and small size classes from Millsap Cache differ significantly in size (RRPP = 10,000, Pr>d = 0.0035), but not shape ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Pairwise comparisons demonstrate that elliptical bifaces from Millsap Cache and Jowell Farm in the large size class are morphologically consistent, but those in the small size class differ in size (RRPP = 10,000, Pr>d = 0.0001) ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Thus, the pattern that emerges is one of differential retouch _intensity_, where isometric reduction strategies of one group (Millsap Cache) may correlate with differing retouch practices that articulate with locally abundant raw material (Kay County flint). This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw material size is insufficient, making it necessary to import the requisite raw material from an extralocal source; in this case, Edwards chert from central Texas.

### Morphological disparity

An analysis of morphological disparity was used to identify potential differences in morphological diversity between size classes \citep{RN11107,RN7041,RN5694}. No difference was found for elliptical biface shape or size by site/raw material, but the comparison of morphological diversity by size class demonstrated that biface shapes in the small size class from Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache. This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more targeted---potentially due to a lack of abundant local raw material---resulting in a broader range of biface shapes in the small size class. Conversely, these results also suggest that Caddo knappers at Millsap Cache employed a more standardised (specialised?) approach to reduction practices.

# Discussion and conclusion

@@

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074). Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by support from the Heritage Research Center at Stephen F. Austin State University.

### 2022-11-01 ms/ms/ms.tex
<edit ms>

\articletype{ARTICLE TEMPLATE}

@@

\author{\name{Robert Z. Selden, Jr.$^{a}$, John E. Dockall$^{b}$, David
H. Dye$^{c}$}
\affil{$^{a}$Heritage Research Center, Stephen F. Austin State
University; Department of Biology, Stephen F. Austin State University;
Texas Archeological Research Laboratory, The University of Texas at
Austin; and Cultural Heritage Department, Jean Monnet
University; $^{b}$Stantec, Inc.; $^{c}$Department of Earth Sciences, The
University of Memphis}
}

\thanks{CONTACT Robert Z. Selden,
Jr.. Email: \href{mailto:zselden@sfasu.edu}{\nolinkurl{zselden@sfasu.edu}}, John
E.
Dockall. Email: \href{mailto:john.dockall@stantec.com}{\nolinkurl{john.dockall@stantec.com}}, David
H.
Dye. Email: \href{mailto:daviddye@memphis.edu}{\nolinkurl{daviddye@memphis.edu}}}

@@

\begin{abstract}
Elliptical bifaces are prevalent at Mississippian sites throughout the
American Southeast. Those from Millsap Cache and Jowell Farm (41AN13)
comprise two of the largest samples of this ill-understood stone tool
from the ancestral Caddo area. Elliptical bifaces from Millsap Cache
were produced using Kay County flint, while those from Jowell Farm were
manufactured using Edwards chert, providing for an empirical test of
morphological differences as a function of raw material. The sample of
elliptical bifaces was divided into two size classes; one conceptually
reflective of production (large), and the other with local reduction
practices (small). Size classes were used to assess whether
modifications by Caddo knappers may have yielded
similar---convergent---biface shape in the small class. Size classes
were also used to test the hypothesis that greater morphological
variation would be apparent in the small class due to idiosyncratic
responses related to local retouch practices. Results demonstrate that
elliptical biface shape does not differ by raw material, but size does.
This suggests that a shared, and consistent, mental template was
maintained independent of biface size. The subsequent analysis of
elliptical biface morphology by size class demonstrated that size does
not differ by raw material in the large class, but in the small class,
it does. This finding supports the argument that elliptical biface
morphology diverges through local reduction practices. As expected,
greater shape diversity occurs in the small class, where Jowell Farm
bifaces were found to be more morphologically diverse than those from
Millsap Cache. Distinct local reduction practices are advanced as the
driver of extant morphological differences found in elliptical bifaces.
\end{abstract}

\begin{keywords}
American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell
knife; Jowell knives; lithics; oblique parallel flaking;
archaeoinformatics; museum studies; digital humanities; non-western art
history; geometric morphometrics; STEM; STEAM
\end{keywords}

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/elliptical.illustration.png}
\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production. The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region.}
\label{fig:illustrated}
\end{figure}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

To assess the variable impacts that Caddo retouch practices may have had
on general morphology, the sample of elliptical bifaces was binned into
two size classes using centroid size. Due to potential morphological
differences associated with the use of two distinct raw materials (Kay
County flint and Edwards chert), the sample was subset by site prior to
calculating the mean centroid size for each. Bifaces larger than the
mean were then assigned to the large category, those smaller than the
mean were assigned to the small category, and the two datasets were then
joined in advance of analysis
(\href{https://seldenlab.github.io/elliptical.bifaces/define-size-classes.html}{Supplementary
Materials, Ch. 1}).

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/ellipticalbifaces.png}
\caption{Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the flaking index developed and advanced by \citet{RN11099}, \citet{RN9242}, and \citet{RN11098}.}
\label{fig:elliptical}
\end{figure}

To identify which edge was most heavily retouched, we employed a
modified approach to the flaking index initially developed by
\citet{RN11099}, and later advanced and refined by \citet{RN9242}, then
\citet{RN11098}. The modified approach uses counts of flake scars from
each edge in the two most heavily worked areas, paired with a measure of
edge length inclusive of curvature, which was calculated using ImageJ
(Abramoff et al.~2004, Rasband 1997, Schneider et al.~2012). The number
of flake scars was then divided by the length of the worked edge, with
the most heavily worked edge identified by the greater value.

@@

The same landmark/semilandmark configuration was used for both analyses,
and the landmarking protocol employs three landmarks; two horizontal
tangents (top/bottom), and the third placed at the furthest extent of
the retouched edge bearing the heaviest amount of retouch (see Figure
\ref{fig:elliptical}). Equidistant semilandmarks were placed between
each landmark, and all landmarks and semilandmarks were applied using R
4.2.1 (R Core Development Team, 2022) and the StereoMorph package (Olsen
and Westneat 2015).

Landmark data were aligned to a global coordinate system (Bookstein et
al.~1999, Gunz et al.~2005, Kendall 1981, 1984, Slice 2001), achieved
through generalised Procrustes superimposition (Bookstein 1986, Rohlf
and Slice 1990, Rohlf 1999) in R using the geomorph (Adams et al.~2017,
Adams and Otarola-Castillo 2013, Baken et al.~2021) and RRPP packages
(Adams and Collyer 2015, Collyer and Adams 2018). Procrustes
superimposition translates and rotates the coordinate data to allow for
comparisons among objects, while also scaling each biface using
unit-centroid size---the square root of the sum of squared distances
from each landmark to the specimen's centroid (Chapman 1990, Dryden and
Mardia 1998, Gower 1975, Rohlf and Slice 1990). The geomorph package
uses a partial Procrustes superimposition that projects the aligned
specimens into tangent space subsequent to alignment in preparation for
the use of multivariate methods that assume linear space (Dryden and
Mardia 1993, Kent and Mardia 2001, Rohlf 1999, Slice 2001).

>>>>

\articletype{ORIGINAL RESEARCH ARTICLE}

@@

\author{\name{Robert Z. Selden, Jr.$^{a}$, John E. Dockall$^{b}$, and
David H. Dye$^{c}$}
\affil{$^{a}$Heritage Research Center, Stephen F. Austin State
University; Department of Biology, Stephen F. Austin State University;
Texas Archeological Research Laboratory, The University of Texas at
Austin; and Cultural Heritage Department, Jean Monnet
University; $^{b}$Stantec, Inc.; $^{c}$Department of Earth Sciences, The
University of Memphis}
}

\thanks{CONTACT Robert Z. Selden,
Jr.. Email: \href{mailto:zselden@sfasu.edu}{\nolinkurl{zselden@sfasu.edu}}, John
E.
Dockall. Email: \href{mailto:john.dockall@stantec.com}{\nolinkurl{john.dockall@stantec.com}}, and
David H.
Dye. Email: \href{mailto:daviddye@memphis.edu}{\nolinkurl{daviddye@memphis.edu}}}

@@

\begin{abstract}
Elliptical bifaces are prevalent at Mississipian sites throughout the
American Southeast. Those from Millsap Cache and Jowell Farm comprise
two of the largest samples of this ill-understood stone tool from the
ancestral Caddo area. Elliptical bifaces from Millsap Cache were
reportedly produced using Kay County flint, while those from Jowell Farm
were manufactured using Edwards chert, providing for an empirical test
of morphological differences as a function of raw material. The sample
of elliptical bifaces was divided into two size classes; one
conceptually reflective of production (large), and the other with local
reduction practices (small). Size classes were used to assess whether
modifications by Caddo knappers may have yielded
similar---convergent---biface shapes in the small class. Size classes
were also used to test the hypothesis that greater morphological
variation would be apparent in the small class due to idiosyncratic
responses related to local retouch practices. Results demonstrate that
elliptical biface shape does not differ by raw material, but size does.
This suggests that a shared, and consistent, mental template was
maintained independent of biface size. The subsequent analysis of
elliptical biface morphology by size class demonstrated that size does
not differ by raw material in the large class, but in the small class,
it does. This finding supports the argument that elliptical biface
morphology diverges through local reduction practices. As expected,
greater shape diversity occurs in the small class, where Jowell Farm
bifaces were found to be more morphologically diverse than those from
Millsap Cache. Distinct local reduction practices are advanced as the
driver of extant morphological differences found in elliptical bifaces.
\end{abstract}

\begin{keywords}
American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell
knife; Jowell knives; Jowell biface; lithics; oblique parallel flaking;
archaeoinformatics; museum studies; digital humanities; non-western art
history; geometric morphometrics; STEM; STEAM
\end{keywords}

@@

\hypertarget{methods-and-results}{%
\section{Methods and results}\label{methods-and-results}}

To assess the variable impacts that Caddo retouch practices may have had
on general morphology, the sample of elliptical bifaces was first binned
into two size classes using centroid size
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Due to potential morphological differences associated with
the use of two distinct raw materials (Kay County flint and Edwards
chert), the sample was subset by site prior to calculating the mean
centroid size for each. Raw material for bifaces from Millsap Cache were
identified by \citet{RN11461} as Kay County flint, and those from Jowell
Farm were identified as Edwards Chert due to their size, the colours
that they fluoresce beneath low and high frequency UV light, and the
paucity of similarly large locally-available raw material sources.
Bifaces larger than the mean were then assigned to the large category,
those smaller than the mean were assigned to the small category, and the
two datasets were then joined in advance of analysis
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}).

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/ellipticalbifaces.png}
\caption{Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area at top right identified using a modified version of the flaking index developed and advanced by \citet{RN11099}, \citet{RN9242}, and \citet{RN11098}.}
\label{fig:elliptical}
\end{figure}

To identify which edge was most heavily retouched, we employed a
modified approach to the flaking index initially developed by
\citet{RN11099}, and later advanced and refined by \citet{RN9242}, then
\citet{RN11098}. The modified approach uses counts of flake scars from
each edge in the two most heavily worked areas, paired with a measure of
edge length inclusive of curvature, which was calculated using ImageJ
\citep{RN11146,RN11147,RN11148}. The number of flake scars was then
divided by the length of the worked edge, with the most heavily worked
edge identified by the greater value.

@@

The same landmark/semilandmark configuration was used for both analyses,
and the landmarking protocol employs three landmarks; two horizontal
tangents (top/bottom), and the third placed at the furthest extent of
the retouched edge bearing the heaviest amount of retouch (see Figure
@ref(fig:elliptical)). Equidistant semilandmarks were placed between
each landmark, and all landmarks and semilandmarks were applied using R
\citep{R} and the \texttt{StereoMorph} package \citep{RN9091}.

\hypertarget{generalised-procrustes-analysis}{%
\subsubsection{Generalised Procrustes
Analysis}\label{generalised-procrustes-analysis}}

Landmark data were aligned to a global coordinate system
\citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through
generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R
using the \texttt{geomorph} and \texttt{RRPP} packages
\citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition
translates and rotates the coordinate data to allow for comparisons
among objects, while also scaling each biface using unit-centroid
size---the square root of the sum of squared distances from each
landmark to the specimen's centroid
\citep{RN11139,RN11140,RN11564,RN478}. The \texttt{geomorph} package
uses a partial Procrustes superimposition that projects the aligned
specimens into tangent space subsequent to alignment in preparation for
the use of multivariate methods that assume linear space (Figure
\ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/gpa.png}
\caption{Results of generalised Procrustes analysis for elliptical bifaces from the Millsap Cache and Jowell Farm. Mean consensus configuration shown in black; individual bifaces in gray.}
\label{fig:gpa}
\end{figure}

\hypertarget{principal-components-analysis}{%
\subsubsection{Principal Components
Analysis}\label{principal-components-analysis}}

Principal components analysis \citep{RN1746} was used to visualise shape
variation among the elliptical bifaces, and the scatterplot and convex
hulls represent the dispersion of shapes in tangent space (Figure
\ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}. The shape space
described by each principal axis is commonly visualized using thin-plate
spline warping of a reference image \citep{RN1731,RN479}.

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/pca.png}
\caption{PCA summarizing shape variation in elliptical bifaces by site and size class. }
\label{fig:pca}
\end{figure}

\hypertarget{procrustes-anova}{%
\subsubsection{Procrustes ANOVA}\label{procrustes-anova}}

To assess whether shape changes with size (allometry), and whether shape
and size differed by raw material and size class, Procrustes ANOVAs
\citep{RN1749} were run that enlist effect-sizes (zscores) computed as
standard deviates of the generated sampling distributions
\citep{RN1756}. A residual randomization permutation procedure (RRPP; n
= 10,000 permutations) was used for all Procrustes ANOVAs
\citep{RN1655,RN11775}, which has higher statistical power and a greater
ability to identify patterns in the data should they be present
\citep{RN1719}.

The test for allometry in the sample did not return a significant
result, suggesting that elliptical biface shape and size may follow a
pattern of isometry, where shape remains constant through the reduction
process
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Elliptical bifaces were also found to express a consistency
of shape across all analyses
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Results demonstrated a significant difference in size by
site/raw material (RRPP = 10,000, Pr(\textgreater F) = 0.0007); however,
the subsequent analysis of elliptical bifaces by size class demonstrates
that the pattern is more readily attributable to reduction
practices/intensity rather than nodule and raw material size
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}).

The analysis of morphology by size class demonstrates a significant
difference in size between size classes; however, that pattern deviated
from our expectations. Elliptical bifaces from Jowell Farm differ in
neither shape nor size between the size classes
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Elliptical bifaces in the large and small size classes from
Millsap Cache differ significantly in size (RRPP = 10,000,
Pr\textgreater d = 0.0035), but not shape
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Pairwise comparisons demonstrate that elliptical bifaces
from Millsap Cache and Jowell Farm in the large size class are
morphologically consistent, but those in the small size class differ in
size (RRPP = 10,000, Pr\textgreater d = 0.0001)
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Thus, the pattern that emerges is one of differential
retouch \emph{intensity}, where isometric reduction strategies of one
group (Millsap Cache) may correlate with differing retouch practices
that articulate with locally abundant raw material (Kay County flint).
This stands in contrast to the comparatively minimal retouch intensity
of another group (Jowell Farm) where local raw material size is
insufficient, making it necessary to import the requisite raw material
from an extralocal source; in this case, Edwards chert from central
Texas.

\hypertarget{morphological-disparity}{%
\subsubsection{Morphological disparity}\label{morphological-disparity}}

An analysis of morphological disparity was used to identify potential
differences in morphological diversity between size classes
\citep{RN11107,RN7041,RN5694}. No difference was found for elliptical
biface shape or size by site/raw material, but the comparison of
morphological diversity by size class demonstrated that biface shapes in
the small size class from Jowell Farm occupy a significantly greater
range of morphospace than those from Millsap Cache. This pattern
suggests that the comparatively minimal retouch intensity associated
with Caddo knappers at Jowell Farm may have been more
targeted---potentially due to a lack of abundant local raw
material---resulting in a broader range of biface shapes in the small
size class. Conversely, these results also suggest that Caddo knappers
at Millsap Cache employed a more standardised (specialised?) approach to
reduction practices.

\hypertarget{discussion-and-conclusion}{%
\section{Discussion and conclusion}\label{discussion-and-conclusion}}
